---
title: "Introduction to truffle"
author: "Ian Hussey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to truffle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

{truffle} is an R package for teaching data processing. It lets you create
datasets with known effects to be rediscovered (*truffles*, via `truffle_`
functions), and then create data-processing headaches that have to be solved
(*dirt*, via `dirt_` functions). Students then search for truffles among the
dirt.

Generated datasets can include demographics and item-level Likert responses.
Known effects (truffles) can be buried in the data, including differences in
sum-score means between conditions, known correlations between outcomes' sum
scores, and known Cronbach's alpha values for each scale. Data can also be made
messy, given impossible values, or given missingness, to create data-processing
challenges.

The package is intentionally narrow: it is designed for a specific teaching use
case and is not (yet) highly flexible. Currently the `truffle_` functions
generate a single design: a between-groups experiment with equal group sizes,
plus a single-group cross-sectional variant.

```{r setup}
library(truffle)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

set.seed(42)
```

# Create a truffle

## Generate clean data

The `truffle_likert()` function generates item-level Likert data for a
between-groups experiment with:

- Two conditions (control vs. treatment, renameable)
- An arbitrary number of outcome scales
- A known Cronbach's alpha for each scale
- A known number of items per scale
- A known number of Likert response options (`1:k`)
- Known correlations between the latent scale scores
- A known *approximate* Cohen's *d* between the two conditions (the recovered
  value differs from the requested one because of reliability, and distortion
  when converting continuous data to Likert)

```{r}
dat_truffle <-
  truffle_likert(study_design = "factorial_between2",
                 n_per_condition = 30,
                 factors  = "X1_latent",
                 prefixes = "X1_item",
                 alpha = .70,
                 n_items = 5,
                 n_levels = 7,
                 approx_d_between_groups = 0.50,
                 seed = 42)

dat_truffle |>
  head() |>
  kable()
```

## Add demographics

```{r}
dat_truffle <- dat_truffle |>
  truffle_demographics()

dat_truffle |>
  head() |>
  kable()
```

## Calculate sum scores by scale

```{r}
dat_truffle_sumscores <- dat_truffle |>
  truffle_sum_scores_by_scale() |>
  select(condition, X1_sum)

# N, M, SD per condition
dat_truffle_sumscores |>
  pivot_longer(cols = starts_with("X"),
               names_to = "scale",
               values_to = "sum_score") |>
  group_by(scale, condition) |>
  summarize(n = n(),
            mean = round(mean(sum_score, na.rm = TRUE), 2),
            sd = round(sd(sum_score, na.rm = TRUE), 2),
            .groups = "drop") |>
  kable()
```

## Check the generated data

`truffle_check()` recovers Cohen's *d* per scale, the correlation matrix among
sum scores, and item-level histograms by condition.

```{r, fig.height = 5}
check <- truffle_check(dat_truffle)

check$cohens_d
check$r
check$item_histograms
```

# Add dirt

Now generate data for the same study, but add dirt step by step.

```{r}
dat <-
  truffle_likert(n_per_condition = 30,
                 factors  = "X1_latent",
                 prefixes = "X1_item",
                 alpha = .70,
                 n_items = 5,
                 n_levels = 7,
                 approx_d_between_groups = 0.50,
                 seed = 42) |>
  # messy demographics with misspellings and erroneous entries
  dirt_demographics() |>
  # impossible values beyond the scale range
  dirt_impossible_values(prop = .04, replacement_value = 8) |>
  # reaction times, then format them with thousands separators and units
  mutate(completion_time = truffle_reaction_times(n = n())) |>
  dirt_numbers(cols = "completion_time") |>
  # mixed date formats
  dirt_dates(col = "date") |>
  # missingness with irregular codings
  dirt_missingness(prop = .05) |>
  # a column holding more than one variable (block and trial)
  dirt_untidy(col = "block_trial") |>
  # duplicate rows
  dirt_duplicates(prop = 0.05) |>
  # move some columns before mangling the names makes it hard
  relocate(date, .after = "id") |>
  relocate(completion_time, .after = "id") |>
  relocate(block_trial, .after = "id") |>
  # poorly named columns that disagree with R
  dirt_colnames()

dat |>
  head() |>
  kable()
```

Header rows can also be prepended, so that reading the file into R requires
skipping them. Here we write to a temporary file and read it back both naively
and correctly.

```{r}
dat_with_header <- dat |>
  dirt_header_rows()

path <- file.path(tempdir(), "dat_truffle_and_dirt.csv")
readr::write_lines(dat_with_header, path)

# naive import: the header rows corrupt the parse
readr::read_csv(path, show_col_types = FALSE) |>
  head(3) |>
  kable()

# correct import: skip the two header rows
readr::read_csv(path, skip = 2, show_col_types = FALSE) |>
  head(3) |>
  kable()
```

# Snuffle functions to help find truffles

`snuffle_` functions help the user find truffles (underlying effects) among the
dirt (data-processing issues).

## Sum scores

Calculating sum scores is a very common task, and yet there are a surprising
number of ways to get it wrong given impossible scores, missingness, and
reversed items. `snuffle_sum_scores()` computes sum scores (optionally imputing
missing values from each participant's mean), excludes out-of-range values, and
returns diagnostics on missing and impossible items. It can also reverse-score
specified items.

```{r}
dat_snuffle <-
  truffle_likert(n_per_condition = 20,
                 factors  = "X1_latent",
                 prefixes = "X1_item",
                 alpha = .70,
                 n_items = 5,
                 n_levels = 7,
                 approx_d_between_groups = 0.50,
                 seed = 42) |>
  truffle_demographics() |>
  dirt_missingness(prop = .10, dirtier = FALSE) |>
  dirt_impossible_values(prop = .04, replacement_value = 8)

dat_snuffle |>
  snuffle_sum_scores("X1_", min = 1, max = 7, id_col = "id") |>
  select(id, condition, X1_sumscore, X1_nonmissing_n, X1_impossible_n) |>
  head(10) |>
  kable()
```

# More complicated truffles

## Multiple correlated outcomes

```{r, fig.height = 8, fig.width = 8}
dat_complex <-
  truffle_likert(n_per_condition = 100,
                 factors  = c("X1_latent", "X2_latent", "X3_latent"),
                 prefixes = c("X1_item", "X2_item", "X3_item"),
                 alpha = c(.70, .75, .80),
                 n_items = c(6, 5, 7),
                 n_levels = 7,
                 r_among_outcomes = 0.50,
                 approx_d_between_groups = c(0.50, 0.20, 1.00),
                 seed = 42)

complex_check <- truffle_check(dat_complex)
complex_check$cohens_d
complex_check$r
```

## Cross-sectional data

The cross-sectional design lets you specify a full latent correlation matrix
among outcomes and recover it from the sum scores.

```{r}
R <- matrix(c( 1, .1, .3,
              .1,  1, .5,
              .3, .5,  1),
            3, 3, byrow = TRUE)

dat_cs <-
  truffle_likert(study_design = "crosssectional",
                 n_per_condition = 500,
                 factors  = c("X1_latent", "X2_latent", "X3_latent"),
                 prefixes = c("X1_item", "X2_item", "X3_item"),
                 alpha = c(.90, .90, .70),
                 n_items = c(10, 7, 15),
                 r_among_outcomes = R,
                 n_levels = c(10, 7, 7),
                 seed = 42) |>
  snuffle_sum_scores("X1_", min = 1, max = 10, id_col = "id") |>
  snuffle_sum_scores("X2_", min = 1, max = 7,  id_col = "id") |>
  snuffle_sum_scores("X3_", min = 1, max = 7,  id_col = "id")

dat_cs |>
  select(X1_sumscore, X2_sumscore, X3_sumscore) |>
  cor() |>
  round(2) |>
  kable()
```
