
{truffle} creates datasets with various known effects to be rediscovered, for teaching purposes. Datasets contain demographics data and item-level Likert responses. Known effects (truffles) can be buried in the data including differences in sum-score means between conditions, known correlations between the different outcomes' sum-scores, known Cronbach's alpha values for each scale, etc. Data can also be made messy, contain impossible values, or contain missingness, to create data processing challenges.

The package's functions are currently quite fragile: it is designed for a specific internal use case in our teaching and not (yet) highly flexible, nor does it contain tests or handle errors or make them visible. Currently it can only generate data for a single design: a between groups experiment with equal sample sizes.

# Dependencies

```{r}

#library(truffle)

library(lavaan)
library(dplyr)
library(tidyr)
library(tidyselect)
library(latent2likert)
library(ggplot2)
library(missMethods)

source("../R/add_demographics_messy.R")
source("../R/add_demographics.R")
source("../R/add_impossible_values.R")
source("../R/add_missingness.R")
source("../R/check_generated_data.R")
source("../R/continuous_to_likert_by_condition.R")
source("../R/generate_data_likert_two_conditions.R")
source("../R/make_lavaan_kfactor_corr.R")
source("../R/sum_scores_by_scale.R")

```

# Generate data

- Item level Likert data (no choice)
- Between subjects experiment (control vs intervention) (2 conditions only, but can be renamed)
- Three outcome variables (arbitrary number)
- Known Cronbach's alpha for each scale (arbitrary number)
- Known number of items per scale (arbitrary number)
- Known number of Likert response options (1:k)
- Known correlations between the latent scale scores
- Known APPROXIMATE Cohen's d between the two conditions' latent scale scores (arbitrary number, but recovered value will differ due to reliability, distortion due to converting continuous data to Likert, etc.)

```{r}

dat_clean <- 
  generate_data_likert_two_conditions(n_per_condition = 50,
                                      factors  = c("X1_latent", "X2_latent", "X3_latent"),
                                      prefixes = c("X1_item", "X2_item", "X3_item"),
                                      alpha = c(.70, .75, .80),
                                      n_items = c(10, 7, 15),
                                      n_levels = 7,
                                      r_among_outcomes = 0.50,
                                      approx_d_between_groups = 0.50,
                                      seed = 42) |>
  add_demographics() 

dat_clean

```

Check that the sum scores conform to the predefined properties and that the item level data is approximately normal.

```{r}
#| fig-height: 20
#| fig-width: 20

check_generated_data(dat_clean)

```

Generate data for the same study but make the demographics data mess, add missingness, and add impossible values to the item level data.

```{r}

dat_messy_missing <- 
  generate_data_likert_two_conditions(n_per_condition = 50,
                                      factors  = c("X1_latent", "X2_latent", "X3_latent"),
                                      prefixes = c("X1_item", "X2_item", "X3_item"),
                                      alpha = c(.70, .75, .80),
                                      n_items = c(10, 7, 15),
                                      n_levels = 7,
                                      r_among_outcomes = 0.50,
                                      approx_d_between_groups = 0.50,
                                      seed = 42) |>
  add_demographics_messy() |>
  add_missingness(proportion_missing = .05) |>
  add_impossible_values(proportion_impossible = .04, replacement_value = 8)

dat_messy_missing

```


